ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2024-12-04 paper/Manuscript-Data-Package.Rmd
Started editing the paper after Bruno's work

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes the open data product `ActiveCA`, with Canadian time use data about active travel. `ActiveCA` is an `R` data package that contains analysis-ready data from cycles  General Social Survey cicles regarding active travel in Canada. The package provides data set detailing active trip origins, destinations, and duration, covering a wide variety of locations, such as home, work or school, libraries, museums, restaurants, bars, sports centers, health clinics, place of worship, and others. The package also details the respondent's region characteristics, specifying wheter they live in a metropolitan area and their province of residency. `ActiveCA` also provides pre-estimated impedance functions for active travel for Canadian Metropolitan Area and Census Agglomerations, which can be used in accessibility analysis to calculate the cost of travel between different locations. The package will continue to expand with contributions from the authors and the broader community through requests in the future. `ActiveCA` is freely accessible for exploration and download from the associated Github repository, where the documentation and code involved in creating and manipulating data are detailed."

keywords: Active; mobility; walking; cycling; travel time; impedance; transportation; R 
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r load-data, include = FALSE}
# Load data
load(paste0(here(), "/data/gss_episodes.Rda"))
load(paste0(here(), "/data/gss_impedances.Rda"))
```

@@

This paper presents the open data product `ActiveCA`. Open data products
(ODPs) are the outcome of a transparent process that transforms raw data
(open or not) into analysis-ready data, in which all stages of
development follow open principles [@arribas-bel2021]. ODPs differ from
general open data due to their high utility, added value and open
availability.

`ActiveCA` is an `R` data package that provides analysis-ready data from the General Social Survey of Canada Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) [@statisticscanada2024] regarding active travel. These data can be used to study walking and cycling in Canada for various purposes over time, and to calculate impedance functions for accessibility analysis. The package provides pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas.

To create this `R` data package, we collected, cleaned and processed the
Time Use collections from the GSS surveys to make them ready for
analysis. An `R` data package contains code, data and documentation in a
standardized format that can be installed by `R` users via a centralized
software repository, such as CRAN (Comprehensive R Archive Network) and
GitHub. Although the GSS surveys are publicly sourced managed by
Statistic Canada [@statisticscanada2024], preparing them for analysis can be time-consuming, tedious and perhaps not even possible for those who try, due to a lack of documentation or prior knowledge.

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. `ActiveCA` is
freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. In addition of enabling to obtain impedance functions, `ActiveCA` can be adopted in various applications that even go beyond the range of possibilities we have imagined. Not only the data, but also all the code documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

@@

Statistics Canada [-@statisticscanada2024] conducts GSS surveys to
obtain data on social trends to track changes in Canadians' living
conditions and well-being over time. This survey is used to understand
how citizens spend and manage their time and what factors contribute to
their happiness and stress. Created in 1985, the survey is part of a
series of independent, annual, and cross-sectional surveys.

In addition to the main topic, each GSS cycle includes new content that
addresses emerging and policy-relevant issues. Every five to seven
years, the Time Use Surveys [@statisticscanada2022] collect data on
respondents' participation and time spent on a wide range of everyday
activities using a 24-hour retrospective diary, with information on the
location of these activities (e.g. at home, at work, etc.) and, for
non-personal activities, the people who were present with the respondent
at the time of the activity. In addition, time-use surveys also cover
topics related to leisure time, work-life balance, health, commuting,
culture and sports, and many others.

The most recent time use survey was carried out in 2022 [@wray2024].
However, the 2022 dataset has not been fully published and, because of
this, our analysis focused on the surveys from 1986 to 2015 (1986, 1992,
1998, 2005, 2010 and 2015). Time Use surveys are composed of two data
sets, the main one and the episode file, explained in the following
subsections.

@@

The main file compiles an large array of aggregated data, summarizing
the answers to the questionnaire and derived variables that summarize
the respondents' time use across different activities, locations, and
social interactions. This file documents the time and duration that
respondents allocate to each activity and location. The Main File
provides a overview of daily routines and social dynamics, not focusing
on individual activity episodes. Additionally, this file categorizes
activities into bigger groups and subcategories, facilitating the data's
analytical utility with additional metrics such as total transit time,
duration spent with household members, and counts of activities and
episodes.

@@

The episode file records detailed data for each activity episode
reported by respondents. The entry includes the start and end times,
duration, location, and accompanying social context, informing when and
where activities occurred and with whom. The file distinguishes itself
by focusing on individual episodes rather than respondents, with the
data structured around the numerous activity instances that compose a
day of the respondent. Although respondent-specific characteristics are
not included within the episode file, it is possible to link the main
file and the episode file by using an identifiable variable present in
both data sets.

# Methodology

For each selected cycle of the GSS surveys, we reviewed the
episode files to identify cases with activities listed as walking or
cycling, selecting the activities immediately before and after the
mobility episode. After that, we labeled the code variables with their appropriate descriptions, identifying each origin and destination, transportation mode, as well the province and urban classification of the respondent's residency.

@@

# Results

@@

`ActiveCA` also enables visual analysis of active travel in Canada using traditional exploratory data analysis techniques. Figures \ref{fig:figure-01} and
\ref{fig:figure-02} show walking and cycling trips from 1992 and 2015
through heat maps. These maps use color gradients to represent the
percentage of trips between various origins and destinations, with
darker colors indicating higher percentages and lighter colors
representing less frequent routes. To avoid overwhelming the reader, we
omitted the heat maps for the other years analyzed.

@@

@@

@@

This article presents `ActiveCA`, an open data product that provides analysis-ready data from Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) of the GSS surveys on active travel in Canada. In the form of an `R` data package, `ActiveCA` was developed after collecting, cleaning, and processing the survey data, providing information on origins, destinations, and duration of active travel, as well other information. Additionally, the package includes a series of pre-estimated impedance functions for walking and cycling trips, considering various destinations and time periods, for Canadian Metropolitan and Census Agglomeration areas.

The value of `ActiveCA` lies in its transparency, accessibility, and
open infrastructure, which facilitates the addition of complementary
data sets in the future. R users can seamlessly explore GSS walking and
cycling episodes along with calibrated impedance functions, with the
option to suggest enhancements to the package as needed. This article
adopts the structure proposed by Anastasia and Páez [-@soukhov2023],
whose work provided essential guidance for the creation of this package.
Similarly, we aim to contribute to the academic community by promoting
transparent research practices that encourage replication and innovation
in related fields. We believe that `ActiveCA` will serve as a basis for
further research on GSS and for the integration of additional data by
the authors or the wider open source community.

@@

The {ActiceCA} R data package can be found and installed on Github
(*link*).

>>>>

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes {ActiveCA}, and open data product with Canadian time use 
data. {ActiveCA} is an `R` data package that contains analysis-ready data related 
to active travel spanning almost 40 years, extracted from cycles 2 (1986), 7 (1992),
12 (1998), 19 (2005), 24 (2010), and 29 (2015) of Canada's General Social Survey.
Active travel is characterized by mode, with walking being part of every cycle 
and bicycling starting in 1992. The attributes of active trips are the types of 
locations of origins and destinations, the duration of trips, and episode weights
for expanding the trips to population-wide estimates. Based on the year of the 
survey, a variety of locations are coded. In earlier cycles, these include home,
work or school, and other's home, whereas in later cycles these are augmented with
locations such as grocery stores, restaurants, outdoor destinations, and others.
The geographical resolution includes the province and whether the episode was in
an urban or rural setting."

keywords: Active; mobility; walking; cycling; travel time; time-use; General Social Survey 
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

<!-- 
FROM THE ABSTRACT:

The package also details the respondent's region characteristics, specifying wheter they live in a metropolitan area and their province of residency. {ActiveCA} also provides pre-estimated impedance functions for active travel for Canadian Metropolitan Area and Census Agglomerations, which can be used in accessibility analysis to calculate the cost of travel between different locations. The package will continue to expand with contributions from the authors and the broader community through requests in the future. {ActiveCA} is freely accessible for exploration and download from the associated Github repository, where the documentation and code involved in creating and manipulating data are detailed.
-->

@@

```{r load-data, include = FALSE}
# Load data
data("gss_episodes")
data("gss_impedances")
```

@@

The objective of this paper is to introduce {ActiveCA}, an open data product with
time use data from Canada's General Social Surveys. Open data products
(ODPs) are the outcome of a process that transforms raw data (open or not) into 
analysis-ready data, following a transparent process in which all stages of
development follow open principles [@arribas-bel2021]. ODPs, while still open, 
differ from general open data in their degree of ease of access, their heightened 
usability, and potentially the value they add to the raw data.

{ActiveCA} provides analysis-ready data concerning active travel in Canada spanning
a period of almost 40 years. The source of these data is Canada's program of 
General Social Surveys (GSSs). This program is designed to provide cross-sectional
data on topics of interest to improve the well-being of Canadians. As part of this
program, every five to seven years the survey is done on the topic of time use. 
Concretely, {ActiveCA} covers Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005),
24 (2010), and 29 (2015) of the GSS. Time use data in these surveys is coded using
a very fine grain, from time spent in chores, leisure, and sleeping, to time spent
working or at school. These surveys have proved valuable in investigations of 
mobility and quality of life [@spinneyTransport2009], the relationship between
active travel and transit use [@lachapelleLonger2016], and travel behavior and time
poverty [@kimFacing2024], to name but a few examples.

For {ActiveCA}, data we extracted all data needed to characterize active travel 
in Canada, namely, episodes where the activity was identified as moving between 
an origin and a destination, either by walking or cycling. Although Statistics 
Canada offers Public Use Microdata Files and documentation for the GSS program 
[see @statisticscanada2024], accessing these files, and preparing  them for analysis
is not a straightforward matter, given their size and complexity.  The  process 
of extracting information of interest from the source files is  time-consuming, 
tedious, and challenging and/or prone to error due to the expertise required to 
work with these files. To create {ActiveCA} we collected, cleaned, and processed
the cycles of the GSS  surveys concerning time use to make them ready for analysis.

{ActiveCA} is distributed as an `R` package with a number of data 
objects and their documentation. `R` packages contain code, data, and documentation
in a standardized format that can be installed by `R` users via a software repository,
such as CRAN (Comprehensive R Archive Network) or GitHub, which makes them an 
adroit medium to distribute analysis-ready data.

Given the level of interest in active travel [e.g., @mccurdySupport2023], reducing
the barriers to using data contained in rich, but difficult to access and use surveys,
such as Canada's GSSs, is a worthy endeavour that can only improve data-driven 
decisions in transportation, urban, and health policy.

<!-- 
is an `R` data package that provides analysis-ready data from the General Social Survey of Canada  [@statisticscanada2024] regarding active travel. These data can be used to study walking and cycling in Canada for various purposes over time, and to calculate impedance functions for accessibility analysis. The package provides pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas.
-->

The rest of this paper discusses the sources of data, and the process implemented
to retrieve and package them. Then, we explain how to use the package and show
some selected examples of analysis to whet the imagination of potential users.
This ODP provides not only data that are easy to use, but also all the code and 
documentation that make this a reproducible research project. In summary, {ActiveCA}
aims to implement and inspire the best principles of open spatial sciences 
[@paez_open_2021; @brunsdon_opening_2021].

@@

Statistics Canada [-@statisticscanada2024] conducts GSS surveys to obtain data 
on social trends to track changes in Canadians' living conditions and well-being
over time. The series of survey on time use are used to understand how Canadian
residents spend and manage their time, and what factors contribute to their 
happiness and stress. The GSS program was created in 1985, and is serialized to 
provide a collection of annual, representative cross-sectional surveys.

The topics of the survey cycle every few years to cover topics that include family,
health, social identity, and every five to seven years time use. The first Canadian
time use survey done as part of the GSS program was conducted in 1986, and the 
most recent was completed in 2015. These Time Use Surveys [@statisticscanada2022]
collect data on respondents' participation and time spent on a wide range of 
everyday activities using a 24-hour retrospective diary, with information on the 
location of these activities (e.g. at home, at work, etc.) and, for non-personal
activities, the people who were present with the respondent at the time of the 
activity. In addition, time-use surveys also cover topics related to leisure time,
work-life balance, health, commuting, culture and sports, and many others.

The Public Use Microdata are released by Statistics Canada in two files: a 
_Main File_ and an _Episode File_. The files are linked by keys that identify 
households, individuals, and episodes (i.e., activities) conducted by individuals.
We discuss these files in more detail in the following section.

@@

The main file of the time use survey compiles a large array of aggregated data, 
summarizing the answers to the questionnaire that describe households and 
individuals, as well as derived variables that summarize the respondents' use of 
time use across different activities, locations, and social interactions. This 
file documents the time and duration that respondents allocate to each activity 
and location. The Main File provides a overview of daily routines and social 
dynamics, not focusing on individual activity episodes. Additionally, this file 
categorizes activities into bigger groups and subcategories, facilitating the 
data's analytical utility with additional metrics such as total transit time, time 
spent with household members, and counts of activities and episodes.

<!-- 
Maybe here provide an example of the structure of this table?
-->

@@

The episode is a much bigger file that records detailed data for each activity 
episode reported by respondents. Each episode represents a single activity and its
duration, and the sum of all episodes throughout the day adds up to 24 hours. Each
entry in this file includes the start and end times of the activity, the duration,
location, and accompanying social context, informing when and where activities 
occurred and with whom. The focus of the Episode File is not on the characteristics
of the respondents, but on the characteristics of the activities, and the data 
are structured around the numerous activity instances that compose a day of the 
respondent. Although respondent-specific characteristics are not included within
the episode file, it is possible to link the main file and the episode file by 
using a key present in both the Main and Episode Files.

<!-- 
Maybe here provide an example of the structure of this table?
-->

# Methods

For each selected cycle of the GSS surveys, we reviewed the episode files to 
identify episodes of movement that involved walking or cycling. This allowed us 
to select the activities immediately before and after the movement episode. After
that, we labeled the code variables with their appropriate descriptions, identifying
each origin and destination, mode of travel, as well the province and urban 
classification of the episode.

<!--

Show a graphical example of this process: how were active trips identified? 
What was the purpose of finding the preceding and succeeding activities?

-->

<!--

Not needed in this paper:

@@

-->

# How to use {ActiveCA}

<!--

The paper does not really need to report "Results". There are too many potential use cases and we would rather let users find their own results depending on what they are interested in. Instead of results, a section on "How to use the package" is appropriate. There are a number of points to highlight here, for example, how to load the package, call the data sets, in particular the use of episode weights to make sure that any statistical analysis is valid.

-->

@@

{ActiveCA} also enables visual analysis of active travel in Canada using traditional 
exploratory data analysis techniques. Figures \ref{fig:figure-01} and \ref{fig:figure-02}
show walking and cycling trips from 1992 and 2015 through heat maps. These maps 
use color gradients to represent the percentage of trips between various origins
and destinations, with darker colors indicating higher percentages and lighter 
colors representing less frequent routes. For conciseness, we omitted the heat maps
for the other years analyzed.

@@

# Python integration

CAN WE PROVIDE PYTHON INTEGRATION? See here:

https://github.com/paezha/idealista18/tree/master/Python

<!--

@@

-->

@@

This paper presents `ActiveCA`, an open data product that provides analysis-ready 
data from Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) 
of the GSS surveys on active travel in Canada. In the form of an `R` data package, 
{ActiveCA} was developed after collecting, cleaning, and processing the survey data, 
providing information on origins, destinations, and duration of active travel, as
well other information.

It is important to remark that the present version of {ActiveCA} covers all Canadian
time use surveys up to 2015. While the most recent time use survey was carried out
in 2022 [@wray2024], the Public Use Microdata Files are currently unavailable, and
at the moment it is estimated that they will only be published in the later part
of 2025. The `R` package will be updated once these files become available.

<!--

Additionally, the package includes a series of pre-estimated impedance functions 
for walking and cycling trips, considering various destinations and time periods, 
for Canadian Metropolitan and Census Agglomeration areas.

-->

The value of {ActiveCA} lies in its transparency, accessibility, and ease of use,
which facilitates the addition of complementary data sets in the future. `R` users
can seamlessly explore GSS walking and cycling episodes <!-- along with calibrated impedance functions-->,
with the option to suggest enhancements to the package as needed. This article adopts
the structure proposed by Anastasia and Páez [-@soukhov2023], whose work provided
essential guidance for the creation of this package. Similarly, we aim to contribute
to the academic community by promoting transparent research practices that encourage
replication and innovation in related fields. We believe that {ActiveCA} will serve
as a basis for further research on GSS and for the integration of additional data by
the authors or the wider open source community.

@@

The {ActiveCA} R data package can be found and installed on Github
(*link*).
